---
title: "Tidy Data"
output: html_notebook
---


# Introduction

A data set is **tidy** if:

- each *variable* is in its own *column*

- each *case* is in its own *row*

- each *value* is in its own *cell*



The steps to follow in a workflow are:

- *Form hypothesis*

- *Design experiment*

- *Collect data*

- Import data into software

- Tidy data into useable form

- *Explore or test*

- Transform the data if required. Do feature engineering

- Write code to apply a modeling algorithm

- Vsualise the data and/or results

- *Communicate results*

- Build app or write paper

- Deploy app or publish paper



# R setup
```{r setup}
ls() # can be used to clear R's memory of any prior stored objects 

library(tidyverse)
#install.packages(babynames)
library(babynames)


# Toy data
cases <- tribble(
  ~Country, ~"2011", ~"2012", ~"2013",
      "FR",    7000,    6900,    7000,
      "DE",    5800,    6000,    6200,
      "US",   15000,   14000,   13000
)
pollution <- tribble(
       ~city, ~size, ~amount,
  "New York", "large",      23,
  "New York", "small",      14,
    "London", "large",      22,
    "London", "small",      16,
   "Beijing", "large",     121,
   "Beijing", "small",     121
)
x <- tribble(
  ~x1, ~x2,
  "A",   1,
  "B",  NA,
  "C",  NA,
  "D",   3,
  "E",  NA
)
# To avoid a distracting detail during class
names(who) <- stringr::str_replace(names(who), "newrel", "new_rel")
```



```{r}
head(cases) # four columns and 3 rows of data. Note that a tibble does not count the row names as a row
# This data is MESSY as we do not have each variable in its own column. eg. country, year and value (i.e. count) are all variables so in a TIDY world there would be a seperate column for each of country, year and value.
```

# Converting data from wide to long format using gather()

## Your Turn 1

On a sheet of paper, draw how the cases data set would look if it had the same values grouped into three columns: **country**, **year**, **cases**

## Your Turn 2

Use `gather()` to reorganize `cases` into three columns: **country**, **year**, and **cases** and save this object as *new.table*.

```{r}
new.table <- gather(cases, key = "year", value = "cases", 2:4) # note have to specify the number of COLUMNS in the data that we want to transform/change using gather()
new.table 
```

Using the Pipe function **%>%** is "neater" than my attempt above. What the function does is to pass the left hand side of the operator to the first argument of the right hand side of the operator. Thus, **iris %>% head()** is equivalent to **head(iris)** 

Often, %>% is called multiple times to "chain" functions together, which accomplishes the same result as nesting. For example the chain: **iris %>% head() %>% summary()** is equivalent to iris being passed to head(), then the result of that passed to summary(). So this is the same as using **summary(head(iris))**.

Some people prefer chaining to nesting because the functions applied can be read from left to right rather than from inside out. Some people also think that chains make for more *beautiful* code.

```{r}
library(dplyr) # using pipes is a "neater" option. Note that this requires the use of additional library dplyr which imports the pipe function from library magrittr

(new.table2 <- cases %>% 
  gather(key = "year", value = "cases", 2:4))
```


Use `gather()` to reorganize `table4a` into three columns: **country**, **year**, and **cases**.

```{r}
table4a # 3 columns and 3 rows
```
  
```{r}
table4a %>%
  gather(key = "year", value = "cases", 2:3, convert = TRUE) # note that convert = TRUE changes year from a character value to an integer value
```

# Converting data from long to wide format using spread()

## Your Turn 3

On a sheet of paper, draw how the pollution data set would look if it had the same values grouped into three columns: **city**, **large**, **small**

```{r}
head(pollution)
```


## Your Turn 4

Use `spread()` to reorganize `table2` into four columns: **country**, **year**, **cases**, and **population**.

```{r}
table2
```


```{r}
spread(table2, key = "type", value = "count", convert = TRUE)
```

Now the column *type* (which consisted of cases and population) has been replaced with two columns of these names, and the values in the *count* column are now in the new *cases* and *population* columns.


Using pipe function this would be coded as:

```{r}
table2 %>%
  spread(key = "type", value = "count", convert = TRUE)
# where: Key = X is the column to use for keys (becomes the new column names); and value = X is the column to use for values (becomes new column cells)
```


# Longer example: Dealing with untidy data using pipe function and both gather() and spread() to pick apart a messy code string used in the data input

## Your Turn 5

The *who* dataset is very messy... the names of column 5 to 60 are a string of information about TB cases, where for example: sp = pulmonary TB, smear positive; sn = pulmonary TB, smear negative etc. The character after this represents if the individual is male (m) or female (f). A code follows this which represents their age where 014 = age group 0 to 14 years; 1524 = age group 15 to 24 years etc.

```{r}
head(who)
```

This information would be simpler to analyse if it were in the long format with the information held in the column code as a single column.

Here we: Gather the 5th through 60th columns of `who` into a key column: value column pair named **codes** and **n**. Then select just the `county`, `year`, `codes` and `n` variables.

```{r}
who %>% # the data we want to manipulate
 gather("codes", "n", 5:60) %>% # use gather function. Specify the new name of the single column we want with the who codes in it as "code", and specify the name of the column we want which will hold the cell values, and then specify the columns that we want to manipulate
 select(-iso2, -iso3) # this removes columns iso2 and iso3 to give us the "neater" dataset
```

## Your Turn 6

We can now Separate the *codes* column into the individual pieces of information that we want. e.g. seperate out the information in the code about sex and age into seperate column of **sex** and **age**.

We do this using seperate() specifying that we need to split the *codes* column into three new columns called "new", "type", "sexage". We can do this using the underscore "_" to identify the seperate parts of the code. 

Note that there are 3 parts to the code: "new", "type of TB such as "sp"", and the "sex & age". In this case we shall call the sex & age group "sexage" for now as they are not seperated by a "_"

```{r}
who %>% 
 gather("codes", "n", 5:60) %>% 
 select(-iso2, -iso3) %>% 
separate(codes, into = c("new", "type", "sexage"), sep = "_") %>%
 select(-new) # this removes the "new" column which does not hold any data of interest for us in this example
```

We can now split the column *sexage* into sex and age by specify the number of *characters* to split by. In this case, we want it to split after ONE character (the m/f character) and put the remaining into a seperate age column.

```{r}
who %>% 
 gather("codes", "n", 5:60) %>%
  separate(codes, into = c("new", "type", "sexage"), sep = "_") %>%
   select(-iso2, -iso3, -new) %>% # to reduce length of code can select all the columns to remove in one go here
  separate(sexage, into = c("sex", "age"), sep = 1) %>%
  drop_na(n) # we can use drop_na to remove na's from the data if desired
```


Sometimes we might want to combine the cells in two seperate columns (or more) into one column to create a code. Lets do this now by re-combining the sex and age to get the "sexage" column back.

```{r}

```



## Your Turn 7

Reshape the layout of this data. Calculate the percent of male (or female) children by year. Then plot the percent over time.

```{r}
# original code
babynames %>%
  group_by(year, sex) %>% 
  summarise(n = sum(n))

# my attempt
babynames %>%
  group_by(year, sex) %>%
  summar
```

***

# Take Aways

Data comes in many formats but R prefers just one: _tidy data_.

A data set is tidy if and only if:

1. Every variable is in its own column
2. Every observation is in its own row
3. Every value is in its own cell (which follows from the above)

What is a variable and an observation may depend on your immediate goal.