---
title: "Tidy Data"
output: html_notebook
---


# Introduction

A data set is **tidy** if:

- each *variable* is in its own *column*

- each *case* is in its own *row*

- each *value* is in its own *cell*



The steps to follow in a workflow are:

- *Form hypothesis*

- *Design experiment*

- *Collect data*

- Import data into software

- Tidy data into useable form

- *Explore or test*

- Transform the data if required. Do feature engineering

- Write code to apply a modeling algorithm

- Vsualise the data and/or results

- *Communicate results*

- Build app or write paper

- Deploy app or publish paper




```{r setup}
ls() # can be used to clear R's memory of any prior stored objects 

library(tidyverse)
#install.packages(babynames)
library(babynames)


# Toy data
cases <- tribble(
  ~Country, ~"2011", ~"2012", ~"2013",
      "FR",    7000,    6900,    7000,
      "DE",    5800,    6000,    6200,
      "US",   15000,   14000,   13000
)
pollution <- tribble(
       ~city, ~size, ~amount,
  "New York", "large",      23,
  "New York", "small",      14,
    "London", "large",      22,
    "London", "small",      16,
   "Beijing", "large",     121,
   "Beijing", "small",     121
)
x <- tribble(
  ~x1, ~x2,
  "A",   1,
  "B",  NA,
  "C",  NA,
  "D",   3,
  "E",  NA
)
# To avoid a distracting detail during class
names(who) <- stringr::str_replace(names(who), "newrel", "new_rel")
```



```{r}
head(cases) # four columns and 3 rows of data. Note that a tibble does not count the row names as a row
# This data is MESSY as we do not have each variable in its own column. eg. country, year and value (i.e. count) are all variables so in a TIDY world there would be a seperate column for each of country, year and value.
```

## Your Turn 1

On a sheet of paper, draw how the cases data set would look if it had the same values grouped into three columns: **country**, **year**, **cases**

## Your Turn 2

Use `gather()` to reorganize `cases` into three columns: **country**, **year**, and **cases** and save this object as *new.table*.

```{r}
new.table <- gather(cases, key = "year", value = "cases", 2:4) # note have to specify the number of COLUMNS in the data that we want to transform/change using gather()
new.table 
```

Using the Pipe function **%>%** is "neater" than my attempt above. What the function does is to pass the left hand side of the operator to the first argument of the right hand side of the operator. Thus, **iris %>% head()** is equivalent to **head(iris)** 

Often, %>% is called multiple times to "chain" functions together, which accomplishes the same result as nesting. For example the chain: **iris %>% head() %>% summary()** is equivalent to iris being passed to head(), then the result of that passed to summary(). So this is the same as using **summary(head(iris))**.

Some people prefer chaining to nesting because the functions applied can be read from left to right rather than from inside out. Some people also think that chains make for more *beautiful* code.

```{r}
library(dplyr) # using pipes is a "neater" option. Note that this requires the use of additional library dplyr which imports the pipe function from library magrittr

(new.table2 <- cases %>% 
  gather(key = "year", value = "cases", 2:4))
```


Use `gather()` to reorganize `table4a` into three columns: **country**, **year**, and **cases**.

```{r}
table4a # 3 columns and 3 rows
```
  
```{r}
table4a %>%
  gather(key = "year", value = "cases", 2:3, convert = TRUE) # note that convert = TRUE changes year from a character value to an integer value
```

## Your Turn 3

On a sheet of paper, draw how this data set would look if it had the same values grouped into three columns: **city**, **large**, **small**

## Your Turn 4

Use `spread()` to reorganize `table2` into four columns: **country**, **year**, **cases**, and **population**.

```{r}
table2
```


```{r}
spread(table2, key = "type", value = "count")
```
Now the column *type* (which consisted of cases and population) has been replaced with two columns of these names.



## Your Turn 5

Gather the 5th through 60th columns of `who` into a key column: value column pair named **codes** and **n**. Then select just the `county`, `year`, `codes` and `n` variables.

```{r}
```

## Your Turn 6

Separate the `sexage` column into **sex** and **age** columns.

```{r}
```

## Your Turn 7

Reshape the layout of this data. Calculate the percent of male (or female) children by year. Then plot the percent over time.

```{r}
# original code
babynames %>%
  group_by(year, sex) %>% 
  summarise(n = sum(n))

# my attempt
babynames %>%
  group_by(year, sex) %>%
  summar
```

***

# Take Aways

Data comes in many formats but R prefers just one: _tidy data_.

A data set is tidy if and only if:

1. Every variable is in its own column
2. Every observation is in its own row
3. Every value is in its own cell (which follows from the above)

What is a variable and an observation may depend on your immediate goal.